# Mark-recapture {#cross}

Mark-recapture is a powerful technique used in ecology and wildlife biology to estimate the population size of a specific animal or species within a given area. This method is based on the concept of sampling, marking, and then recapturing individuals from the target population. By utilizing this approach, researchers can extrapolate the population size and gain valuable insights into population dynamics, species abundance, and conservation efforts.

The fundamental principle behind mark-recapture lies in the idea that the probability of capturing an individual within the population is the same for each capture event. Once an individual is captured, it is marked in some way, making it distinguishable from other members of the population. The marked individuals are then released back into the wild, allowing them to mix freely with the unmarked individuals. Subsequently, during the recapture phase, researchers perform another sampling effort, capturing a new group of individuals from the population. The proportion of marked individuals in the recaptured sample allows scientists to estimate the total population size based on the assumption of a closed population.

In this chapter, we explore the principles and applications of mark-recapture methods, focusing on the Lincoln-Peterson MR estimator in closed populations and the importance of simulation studies to evaluate method performance.

## 2.1 Closed populations

The Lincoln-Peterson MR estimator is one of the most classic and widely used methods for estimating the population size of closed populations. Proposed independently by Peterson in the 1890s and by Lincoln in the 1920s, this estimator relies on two sampling events: the initial capture, where a subset of the population is marked, and the subsequent recapture, where the marked individuals are identified among the newly captured individuals.

The Chapman mark-recapture estimator is a modified version of the traditional Lincoln-Peterson estimator with less bias and is less affected by zeros.

::: {#tri .theorem}
**Chapman MR estimator (1951)** $$\hat{N}_{CPM} = \frac{(n_{1} + 1)(n_{2} + 1)}{m_{2}+1}-1$$

where:

-   $\hat{N}_{CPM}$ is the estimated population size.

-   $n_{1}$ is the number of individuals captured and marked in the first occasion.

-   $n_{2}$ is the number of marked individuals recaptured in the second occasion.

-   $m_{2}$ is the number of recaptured individuals that were already marked in the first occasion.
:::

The accuracy of this estimate relies on a number of assumptions being met.

| **Assumption 1.** The population is closed to additions and deletions;

In other words, no individuals are added to or removed from the population through birth, death, immigration, or emigration during the study period. An exception is when there are equal proportions of additions or deletions for marked and unmarked individuals during the sampling period. In cases where significant movements occur, potential bias might need to be introduced to the results or more advanced models might be needed to account for population dynamics.

| **Assumption 2.** All animals are equally likely to be captured in each sample;

Sometimes after initial capture, animals may learn to avoid the traps or area where the capture occurred, causing them to be captured less frequently ("trap shy"). The opposite ("trap happy") may also occur, making them more likely to be captured. Individuals may also inherently different capture probabilities, and if ignored, the resulting abundance estimates are likely to be negatively biased.

| **Assumption 3.** Marks are not lost and not overlooked by the observer

It is possible that marks may fade, fall off, or be lost over time, making individuals indistinguishable from unmarked ones during recapture events. Ensuring durable and effective marking techniques is essential to maintain the integrity of the mark-recapture process and achieve unbiased estimates.

## 2.2 Simulation study

Simulation studies have become indispensable in ecology and conservation research, providing a controlled and repeatable approach to evaluate the performance of estimation methods. In this section, we will present the results of a simulation study that aims to evaluate the performance of the Lincoln-Peterson MR estimator in closed populations. We will create artificial datasets with with known population sizes, detectability patterns to assess the accuracy and precision of estimators under different scenarios.

For each scenario we simulate 999 surveys, with each survey having a simulated sample size $N$ = 500.

#### Scenario 1: When the underlying assumptions are met

We first looked into the performance of mark-recapture estimator when all the above assumptions are met. Specifically, we will create artificial datasets to mimic a closed population, equal probability of detection of marked and unmarked individuals ($p$ = 0.4), and random sampling. For each replication, we will compare the estimated population size to the true population size ($N$ = 500). We then calculate the mean estimate and the variance of the estimates across replications and thus check if true population size is within the confidence interval of the estimates.

```{r MR1}
sim.mr <- function(n, equal.p, param){
  #browser()
  ## 1. generate n uniform points 
  df <- data.frame(id=rep(1:n, each = 2), obs=rep(1:2, n),
                   x=NA, y=NA, distance=NA, detect=NA)
  
  df$x[df$obs==1] <- runif(n, 0, 2)
  df$y[df$obs==1] <- runif(n, 0, 2)
  df$distance[df$obs==1] <- df$x[df$obs==1] - 1
  
  df$x[df$obs==2] <- runif(n, 0, 2)
  df$y[df$obs==2] <- runif(n, 0, 2)
  df$distance[df$obs==2] <- df$x[df$obs==2] - 1
  
  
  ## 2. detection probability
  if (equal.p == TRUE){
    p1 <- p2 <- param
  }else{
    p1 <- exp(-abs(df$distance[df$obs==1])^2/(2*param^2))
    p2 <- exp(-abs(df$distance[df$obs==2])^2/(2*param^2))
  }
  
  
  ## 3. Mark-Recapture
  df$detect[df$obs==1] <- rbinom(n, 1, p1)  # first occasion
  df$detect[df$obs==2] <- rbinom(n, 1, p2)  # second occasion
  
  
  ## 4. return
  S1 <- nrow(df[df$obs==1 & df$detect==1, ])  # caught by first occasion
  S2 <- nrow(df[df$obs==2 & df$detect==1, ])  # caught by second occasion
  B <- df$detect[df$obs==1]==1 & df$detect[df$obs==2]==1  
  B <- length(B[B==TRUE])  # caught by both occasions
  N.hat <- (S1+1)*(S2+1)/(B+1)-1  # abundance estimate
  return(N.hat)
}

boot.mr <- function(n, equal.p, param){
  input <- rep(n, 999)
  est.N <- unlist(lapply(input, sim.mr, equal.p, param))
  return(est.N)
}
```

```{r}
est.N <- boot.mr(500, TRUE, 0.4)  # bootstrap estimates
ci <- mean(est.N) + c(-1, 1)*1.96*sqrt(var(est.N)/999)  # confidence interval
bias <- mean((est.N-500)/500*100)  # mean percentage relative bias
ci; bias
```

The result shows that Chapman estimator of abundance is nearly unbiased with confidence interval of estimates being [498, 502] and mean percentage relative bias equals to -0.027%.

#### Scenario 2: when inherent capture probabilities are different

In this scenario, we aim to examine the impact of deviations from equal detectability assumption on the accuracy and bias of mark-recapture method. We make probability of detection falls off with distance from the transect line following half-normal distribution in both occasion. Different half-normal parameters are fitted to investigate how they affect the estimate. As shown in the graph below, when detectability decreases more rapidly with increasing distance ($\sigma$ = 0.3), the estimate are more negatively biased. In contrast, if there has less difference between detection probability on the transect and deetction probability on the truncation ($\sigma$ = 0.7), then abundance estimate are less negatively biased.

```{r MR2}

plot1.1 <- function(n){
  #browser()
  par(mfrow=c(1,2))
  
  ## plot 1: half-normal
  x <- seq(0, 1, 0.01)
  y1 <- exp(-x^2/(2*0.3^2))
  y2 <- exp(-x^2/(2*0.5^2))
  y3 <- exp(-x^2/(2*0.7^2))
  plot(x, y1, main = "Half normal", xlab = "Distance", ylab = "Detection probability",
       type="l", lwd=1.5, col="red", xlim=c(0, 1))
  lines(x, y2, col="blue", lwd=1.5, lty=2)
  lines(x, y3, col="darkgreen", lwd=1.5, lty=3)
  legend(x = "topright", legend = c("sigma = 0.3", "sigma = 0.5", "sigma = 0.7"),  
         lty = c(1, 2, 3), col = c("red", "blue", "darkgreen"), lwd = 1.5, cex=0.5) 
  
  
  ## plot 2: bootstrap abundance estimates
  N1 <- boot.mr(n, FALSE, 0.3)
  N2 <- boot.mr(n, FALSE, 0.5)
  N3 <- boot.mr(n, FALSE, 0.7)
  plot(0.3, mean(N1), pch=7, col="red", xlim=c(0.2, 0.8), ylim=c(min(N1), n+10), xlab="sigma", ylab="estimates")
  title(main="Abundance estimates for different detetction functions, N = 500", cex.main=1)
  points(0.5, mean(N2), pch=7, col="blue")
  points(0.7, mean(N3), pch=7, col="darkgreen")
  abline(h=n, lwd=1.5)
  x <- seq(0.3, 0.7, 0.02)
  y <- NULL
  for (i in x){
    meanN <- mean(boot.mr(n, FALSE, i))
    y <- c(y, meanN)}
  lines(x, y, lty=3, lwd=1.5)
  legend(x = "bottomright", legend = c("sigma = 0.3", "sigma = 0.5", "sigma = 0.7"),  
         pch = c(7, 7, 7), col = c("red", "blue", "darkgreen"), cex=0.5) 
  
  
  ## table: 
  bias <- unlist(lapply(c(N1, N2, N3), FUN = function(x) mean((x-500)/500*100)))
  coverage <- NA
  
}
```

```{r}
# I may change the second plot to box plot
# or add a table presenting bias&coverage prob for each sigma
plot1.1(chapman.mr, 500)  # Chapman MR estimator, N
```

#### Limitations of mark-recapture

The validity and accuracy of mark-recapture method relies on critical assumptions. As shown in the simulation above, violation of equal capture probability assumption can lead to biased abundance estimates. Yet, the method has other limitations that may affect result reliability. Small sample size, for example, can lead to decreased precision, larger confidence intervals, and less informative results. Imperfect matching is another circumstance which cause bias in population size estimates. We will investigate more about mismatching case in the following chapters of mark-recapture distance sampling (MRDS) method.
