# 2D Distance Sampling

```{r 2D-setup, include=FALSE, warning = FALSE}
source("code.R")
b = c(4.89553757, 0.03561984); logphi = c(6.61348904, 4.84269089)
w = 1600; ystart = 1300
Nanimals = 200
match_limits = c(300,1000); match_param = c(450,5)
```

## Introduction

2D Distance Sampling builds on standard distance sampling by including the forward distance to the detected animal as well as the perpendicular distance. This removes the assumptions that the animals are uniformly distributed and that p(0) = 1, so this method is useful for situations where the other methods covered previously are not suitable. In particular, 2D distance sampling only requires the transect to be surveyed by one observer. This is not only more efficient but removes the issue of imperfect matching in mark-recapture methods.

Animal movement is also less problematic for 2D distance sampling. Like in standard distance sampling, as long as the movement is slow relative to the speed of the observers it will only cause a small amount of bias. Responsive movement can be accounted for in the animal distribution model so does not cause bias.

The likelihood equation for this model is: $$
L(N, \boldsymbol{\beta}) = {N\choose n}[1-p.(\boldsymbol{\beta},\boldsymbol{\phi})]^{N-n}\prod_{i = 1}^{n}\pi(x_i; \boldsymbol{\phi})S(t_i, x_i; \boldsymbol{\beta}) h(t_i, x_i; \boldsymbol{\beta}) 
$$ where:

\- $n$ is the number of animals observed and $N$ is the total in the population

\- $\pi(x_i; \boldsymbol{\phi})$ is the probability animal i is at perpendicular distance $x_i$. We will refer to $\pi(x_i; \boldsymbol{\phi})$ as the animal distribution function.

\- $h(t_i, x_i; \boldsymbol{\beta})$ is the probability animal i is detected at forward distance $y_i$ and perpendicular distance $x_i$. We will refer to $h(t_i, x_i; \boldsymbol{\beta})$ as the detection function.

\- $S(t_i, x_i; \boldsymbol{\beta}) = e^{-\int_{0}^{t_i}h(u,x;\boldsymbol{\beta})du}$ is the survivor function, i.e. the probability that animal i was undetected until time $t_i$

\- $p.(\boldsymbol{\beta},\boldsymbol{\phi})$ is the average probability of being detected in the region

In practice the conditional likelihood $L(\boldsymbol{\beta}|n)$ is used to calculate the maximum likelihood estimators. It is the same as the equation above, but without the leading binomial term, and so avoids considering the unseen animals. $\hat N$ is then estimated using a Horvitz-Thompson like estimator. The forward distance to the detected animal $y_i$ is used instead of the time $t_i$.

## Simulation: Non Uniform Distribution

We start by simulating attraction to the line by generating 200 animals using a half normal distribution with parameters `r exp(logphi)`. We used a inverse power detection function with parameters `r b`. The parameters chosen are the best model for the real snow leopard prey data.

```{r 2D-simulation, echo = F, warning = F, cache = TRUE}
#suppressWarnings(simdata <- MCR_2D(99,Nanimals, b=b, logphi = logphi, w= w, ystart = ystart, match_limits = match_limits, match_param = match_param))
load("simdata.Rdata")
bias_2d <- 100*(simdata$X2D - Nanimals)/Nanimals
```

We then fitted a model using a half normal distribution for animal density and inverse power detection function. We did not fit any other models for efficiency purposes. This gives a bias of `r mean(bias_2d)` $\pm$ `r 1.96*sqrt(var(bias_2d)/length(bias_2d))`, so we can confirm it is unbiased.

## Simulation: Comparison to Mark Recapture

Double observer mark recapture methods using the Chapman estimator are the current standard for estimating abundance of snow leopard prey. Our previous simulations have shown these have serious flaws, so we used both a 2D distance model and the Chapman estimator to estimate the abundance from the same data (simulated using the same parameters as above).

```{r 2D-histograms, echo = F, fig.cap="Histograms of abundance estimates using 2D distance sampling and the Chapman estimator"}
par(mfrow = c(1,2))
hist(simdata$X2D, main = "2D Distance", xlab = "Estimate of N")
abline(v = Nanimals, col = "blue")
hist(simdata$MCR.nomvmnt, main = "Chapman", xlab = "Estimate of N", xlim = c(min(simdata$MCR.nomvmnt),max(c(max(simdata$MCR.nomvmnt),Nanimals))))
abline(v = Nanimals, col = "blue")
```

The Chapman estimator was significantly negatively biased due to the unmodelled heterogeneity in the population. Adding random animal movement and imperfect matching gives a similar problem.

```{r imperfect-matching-histograms, echo = F, fig.cap = "Histograms of abundance estimates using the Chapman estimator with random movment"}
par(mfrow = c(1,2))
hist(simdata$MCR.move, main = "Perfect Matching", xlab = "Estimate of N", xlim = c(min(simdata$MCR.move),max(c(max(simdata$MCR.move)+10,Nanimals))))
abline(v = Nanimals, col = "blue")
hist(simdata$MCR.imperf, main = "Imperfect Matching", xlab = "Estimate of N", xlim = c(min(simdata$MCR.imperf), max(c(max(simdata$MCR.imperf)+10,Nanimals))))
abline(v = Nanimals, col = "blue")
```

```{r coverage-probabilities, include=FALSE}
cis_2D <- simdata[!is.na(simdata$X2D.lower)&!is.na(simdata$X2D.lower), 2:3]
covprob2D <- 100*length(cis_2D[cis_2D$X2D.lower <= Nanimals &cis_2D$X2D.upper >= Nanimals])/length(cis_2D)
cp.nomvmnt <- 100*length(simdata[simdata$nomvmnt.lower <= Nanimals & simdata$nomvmnt.upper >= Nanimals])/length(simdata$MCR.nomvmnt)
if(length(dim(cis_2D)[1] == 0)){
  print("Hessian matrices could not be computed for any 2D models")
  covprob2D <- NA}
cp.mvmnt <- 100*length(simdata[simdata$mvmnt.lower <= Nanimals & simdata$mvmnt.upper >= Nanimals,])/length(simdata$MCR.move)
cp.imperf <- 100*length(simdata[simdata$imperf.lower <= Nanimals & simdata$nimperf.upper >= Nanimals,])/length(simdata$MCR.imperf)
kable(data.frame("2D" = covprob2D,"nm" = cp.nomvmnt, "m" = cp.mvmnt,"i" = cp.imperf), col.names = c("2D", "Chapman without movement", "Chapman with movement", "Chapman with imperfect matching"), caption = "Coverage Probabilities")
```

From this it is clear that double-observer mark-recapture will yield significantly biased estimates of abundance of snow leopard prey, even if the animals do not move and the observers can perfectly match the individuals seen. We recommend instead using 2D distance sampling, as it accounts for the non uniform animal distribution as well as the declining detectability with distance. It is also requires less time to be spent surveying, as each transect only need to be walked once.




## Further Reading

Borchers, D.L. and Cox, M.J. (2016). Distance sampling detection functions: 2D or not 2D? Biometrics, 73(2), pp.593--602. <doi:https://doi.org/10.1111/biom.12581>.

Borchers, D.L., Buckland, S.T. and Zucchini, W. (2002). Estimating Animal Abundance. [online] Springer Science & Business Media.

Suryawanshi, K.R., Bhatnagar, Y.V. and Mishra, C. (2012). Standardizing the double-observer survey method for estimating mountain ungulate prey of the endangered snow leopard. Oecologia, 169(3), pp.581--590. <doi:https://doi.org/10.1007/s00442-011-2237-0>.
