# Simulation

Simulation studies have become indispensable in ecology and conservation research, providing a controlled and repeatable approach to evaluate the performance of estimation methods. In this study, we aimed to assess how each method performs using a range of scenarios with varying degrees of assumption compliance and violations. In each scenario, artificial datasets are created with with known population sizes, density patterns, detectability patterns, etc. to test the accuracy and precision of estimators.

## Simulation scenarios

### Initial perpendicular distance distribution

Perpendicular distance refers to the minimum distance to the transect line at which animals are detected. 3 different levels are considered: uniformly distributed, avoiding the line and attracted to the line.

```{r initial perp. dist. distribution}
par(mfrow=c(1,3))
xs <- seq(0,1,length=100)

# uniformly distributed
plot(xs, pi.const(x=xs,logphi=NA,w=1),
     type='l',xlab='Perp. distance, x',ylab=expression(pi(x)))

# avoiding the line
plot(xs, pi.chnorm(x=xs,logphi=c(0, -0.5),w=1),
     type='l',xlab='Perp. distance, x',ylab=expression(pi(x)))

# attracted to the line
plot(xs,pi.norm(x=xs,logphi=c(0, 1),w=1),
     type='l',xlab='Perp. distance, x',ylab=expression(pi(x)))

```

```{r distribution-plots, echo = F, fig.cap = "Potential Animal Distributions"}
par(mfrow = c(1,3))
const <- simpop2DLT(600,2000,pi.const, En = 400, fixed.n = T)
plot(const$x, const$y, xlab = "x", ylab = "y", main = "Uniformly Distributed")
abline(v = 0, col = 'red')
avoid <- simpop2DLT(600,2000,pi.chnorm, logphi = c(0,6), En = 400, fixed.n = T)
plot(avoid$x, avoid$y, xlab = "x", ylab = "y", main = "Avoiding Line")
abline(v = 0, col = 'red')
attract <- simpop2DLT(600,2000,pi.hnorm, logphi = 6.5, En = 400, fixed.n = T)
plot(attract$x, attract$y, xlab = "x", ylab = "y", main = "Attracted to Line")
abline(v = 0, col = 'red')
```

### Movement after first detection

For the mark recapture methods there are two detection occasions, which could be , for example, two observers on different decks of a boat or , as in the snow leopard prey data, two observers walking the transect at different times. Animal movement is not an issue for the first scenario as the animals move slowly in comparison to the boat, so will not move far enough to have different detection probabilities between the two observers. However, in our snow leopard prey example, there is 30 minutes between the first and second observer walking the transect so the animals could move significantly between the two observers. We consider two types of movement after the first detection: random movement and avoidant movement away from the transect. We based our simulation on Ibex movement data in Johansson *et al*. (2022), which shows that the distance moved follows roughly a log normal distribution with mean log(12) + 1.5 and standard deviation 1.5,

```{r movement, echo = F, fig.asp= 1.25, warning = F}
library(plotly)
par(mfrow = c(1,2))
sim.dat <- function(n, level){
  #browser()
  ## 1. object positions
  df <- data.frame(obs=rep(1:2, n),
                   x=NA, y=NA, size=NA, level=NA)
  
  df$x[df$obs==1] <- runif(n, -1600, 1600); df$y[df$obs==1] <- runif(n, 0, 600)
  df$size <- rep(sample(4:10, n, replace = TRUE), each=2)
  perp.dist <- abs(df$x[df$obs==1])
  
  
  if (level==1){
    angle <- rwrappedcauchy(n, mu = circular(0), rho = 0, control.circular=list(units="radian"))
    dist <- rlnorm(n,log(12) + 1.5,1.5)
    dist[dist > 750] <- 750
    df$x[df$obs==2] <- df$x[df$obs==1] + dist*cos(angle)
    df$y[df$obs==2] <- df$y[df$obs==1] + dist*sin(angle)/1000
    df$level <- rep("Random", 2*n)
    }else{
    nleft <- nrow(df[df$x <= 0 & df$obs==1, ])
    nright <- nrow(df[df$x > 0 & df$obs==1, ])
    
    angleleft <- rwrappedcauchy(nleft, mu=circular(pi), rho=0.9, control.circular=list(units="radian"))
    df$angle[df$x <= 1 & df$obs==1] <- as.numeric(angleleft)
    angleright <- rwrappedcauchy(nright, mu=circular(0), rho=0.9, control.circular=list(units="radian"))
    df$angle[df$x > 1 & df$obs==1] <- as.numeric(angleright)
    
    dist <- (max(abs(df$x[df$obs==1]))-abs(df$x[df$obs==1]))*runif(n, 0, 0.1)*2.5
    df$x[df$obs==2] <- df$x[df$obs==1] + dist * cos(df$angle[df$obs==1])
    df$y[df$obs==2] <- df$y[df$obs==1] + dist * sin(df$angle[df$obs==1])/1000
    df$level <- rep("Avoidance", 2*n)
    df[ , c("angle")] <- list(NULL)
  }
  
  
  ## 3. return
  return(df)
}

df1 <- sim.dat(50,1)
fig <- df1 %>%
  plot_ly(
    x = ~x,
    y = ~y,
    size = ~size,
    color = ~level,
    frame = ~obs,
    type = 'scatter',
    mode = 'markers')
fig
```

```{r movement-animations}
df2 <- sim.dat(50,2)
fig2 <- df2 %>%
  plot_ly(
    x = ~x,
    y = ~y,
    size = ~size,
    color = ~level,
    frame = ~obs,
    type = 'scatter',
    mode = 'markers')
fig2
```

### Imperfect Matching

Mark recapture methods also assume that the observers can perfectly identify which animals were seen by both observers and which animals were seen by just one observer. This becomes problematic for animals without unique identifying markings in surveys where they could move between detection occasions.

We simulated imperfect matching by setting any animals detected on the second occasion within 300m of an animal detected on the first animal were recorded as the same animal, taking the closest as the match if there were multiple within this range. Any animals further than 1000m were recorded as different, and the probability of matching animals within the range was generated using a inverse power function with parameters $\beta_1$ = 6, $\beta_2$ = 0.000005. (Thanks to the snow leopard prey team for suggesting these limits)

```{r imperfect-matching plot, echo = F}
x <- 0:1000
p <- p.approx(ys <- seq(0, 1300, length.out=100),x , ip0, b=c(6, 0.000005), what = "px")
plot(x,p,ylim = c(0,1), type = "l", xlab = "Distance", ylab = "Probability")
abline(v = 300)
```

## Simulation Results

```{r Sc.1, message=FALSE, warning=FALSE, cache=TRUE}
# Scenario 1
# Initial density: uniform + Movement: avoidance + Matching: perfect
sc1 <- simulation(n=600, b=99, 0, 0, FALSE)
```

```{r Sc.2, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 2
# Initial density: uniform + Movement: avoidance + Matching: imperfect
kable(sc2 <- simulation(n=600, b=9, 0, 0, TRUE))
```

```{r Sc.3, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 3
# Initial density: uniform + Movement: random + Matching: perfect
kable(sc3 <- simulation(n=600, b=9, 0, 1, FALSE))
```

```{r Sc.4, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 4
# Initial density: uniform + Movement: random + Matching: imperfect
kable(sc4 <- simulation(n=600, b=9, 0, 1, TRUE))
```

```{r Sc.5, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 5
# Initial density: avoiding + Movement: avoidance + Matching: perfect
kable(sc5 <- simulation(n=1200, b=9, 1, 0, FALSE))
```

```{r Sc.6, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 6
# Initial density: avoiding + Movement: avoidance + Matching: imperfect
kable(sc6 <- simulation(n=1200, b=9, 1, 0, TRUE))
```

```{r Sc.7, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 7
# Initial density: avoiding + Movement: random + Matching: perfect
kable(sc7 <- simulation(n=1200, b=9, 1, 1, FALSE))
```

```{r Sc.8, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 8
# Initial density: avoiding + Movement: random + Matching: imperfect
kable(sc8 <- simulation(n=1200, b=9, 1, 1, TRUE))
```

```{r Sc.9, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 9
# Initial density: attracted + Movement: avoidance + Matching: perfect
kable(sc9 <- simulation(n=400, b=9, 2, 0, FALSE))
```

```{r Sc.10, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 10
# Initial density: attracted + Movement: avoidance + Matching: imperfect
kable(sc10 <- simulation(n=400, b=9, 2, 0, TRUE))
```

```{r Sc.11, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 11
# Initial density: attracted + Movement: random + Matching: perfect
kable(sc11 <- simulation(n=400, b=9, 2, 1, FALSE))
```

```{r Sc.12, message=FALSE, warning=FALSE, cache = TRUE}
# Scenario 12
# Initial density: attracted + Movement: random + Matching: imperfect
kable(sc12 <- simulation(n=400, b=9, 2, 1, TRUE))
```

## Further Reading

Johansson, Ö., Mishra, C., Chapron, G., Samelius, G., Lkhagvajav, P., McCarthy, T. and Low, M. (2022). Seasonal variation in daily activity patterns of snow leopards and their prey. Scientific Reports, [online] 12(1), p.21681. <doi:https://doi.org/10.1038/s41598-022-26358-w>.
